---
title: "example_confints"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{article_figures}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(WWdPCR)
```

In this document, we compute a couple examples to show how assuming independent binomials for the counts of a drop-off duplex dPCR assay leads to aberrant confidence interval, and compare with using a proper multinomial likelihood.

## Comparison of likelihood surfaces and confidence regions

We plot below the comparison between likelihood surfaces of the product-of-binomial model and the multinomial model, for a special case of $x_0 = 10000, x_1 = 10, x_2 = 100$. We show contours of the confidence regions for $\alpha=0.05, \alpha=0.01, \alpha=0.001$. Notice that in the the product-of-binomial model, the contours exit the [0,1] range for the proportion of variant. 

```{r, fig.dim=c(6,6)}
# log-likelihood based on multinomial distribution derived above
loglik <- function(x_0, x_1, x_2, r, lambda){
    -lambda*x_0 + x_1*log(exp(-lambda*r) - exp(-lambda)) + x_2*log(1 - exp(-lambda*r))
}
# log-likelihood based on multinomial distribution derived above
loglik_ind <- function(x_0, x_1, x_2, r, lambda){
    # log-likelihood based on product of two independent binomial distribution
    n <- x_0 + x_1 + x_2
    p1 <- dbinom(x_2, n, r*lambda, log=T)
    p2 <- dbinom(x_1+x_2, n, lambda, log=T)
    p1+p2
}

# function to make standard errors with delta method from two binomials
se2_r_1 <- function(p1, p2, n1, n2){
    p1/(n1*(1-p1)*log(1-p2)^2) + p2*log(1-p1)^2/(n2*(1-p2)*log(1-p2)^4)
}

x0 <- 10000
x1 <- 10
x2 <- 100 

l_mle = -log(x0/(x0+x1+x2))
r_mle = -log((x0+x1)/(x0+x1+x2))/l_mle

# se2 <- se2_r_1(mle*lam_ml, lam_ml, x0+x1+x2, x0+x1+x2)

xxs <- seq(0,1.5,length.out=500)
yys <- seq(0,0.025,length.out=500)
grid1 <- outer(xxs, yys, function(x,y){loglik(x0, x1, x2, x, y)})
grid2 <- outer(xxs, yys, function(x,y){loglik_ind(x0, x1, x2, x, y)})


par(mfrow=c(1,1))
filled.contour(xxs, yys, grid1, xlab=expression(r), ylab=expression(lambda),
               main="log-likelihood based on\nmultinomial model",
              plot.axes={
                  points(r_mle,l_mle, pch=4, lwd=3)
                  text(r_mle-0.08, l_mle+0.08, "MLE")
                  contour(xxs, yys, grid1, levels=loglik(x0, x1, x2, r_mle, l_mle)-qchisq(c(0.95,0.99,0.999),1)/2,
                          add=T, drawlabels=F)
                  axis(side=1, at=seq(0,1.5,by=0.25)) 
                  axis(side=2, at=seq(0,0.05,by=round(0.025/3,2)))
              })

filled.contour(xxs, yys, grid2, xlab=expression(r), ylab=expression(lambda),
               main="log-likelihood based on\nproduct of binomials model",
              plot.axes={
                  points(r_mle,l_mle, pch=4, lwd=3)
                  text(r_mle-0.08, l_mle+0.08, "MLE")
                  contour(xxs, yys, grid2, levels=loglik_ind(x0, x1, x2, r_mle, l_mle)-qchisq(c(0.95,0.99,0.999),1)/2,
                          add=T, drawlabels=F)
                axis(side=1, at=seq(0,1.5,by=0.25)) 
                axis(side=2, at=seq(0,0.05,by=round(0.025/3,2)))
              })

```

## Comparison of confidence intervals for $\hat r$ in typical situations.

We plot below the (estimated) relative profile log-likelihoods based on the multinomial (red/black) and product of binomial (blue) models, for different values of $x_0, x_1, x_2$. We compare the confidence intervals for $\hat r$ obtained by inverting profile likelihoods ratio tests for these two models, and the confidence intervals produced by the delta method (green). 

```{r, fig.dim=c(8,10), warning=FALSE}

# profile log-likelihood in the multinomial model
lp <- function(x_0, x_1, x_2, r){
    -1*unlist(lapply(r, function(r_temp){
        optim(par=list(l=-log(x_0/(x_0+x_1+x_2))),
          function(x){-loglik(x_0, x_1, x_2, r_temp, x)})$value
    }))   
}

# estimated profile log-likelihood in the multinomial model
le <- function(x_0, x_1, x_2, r){
    x_0*log(x_0/(x_0 + x_1 + x_2)) +
    x_1*log(-x_0/(x_0 + x_1 + x_2) + 
            exp(r*log(x_0/(x_0 + x_1 + x_2)))) + 
    x_2*log(1 - exp(r*log(x_0/(x_0 + x_1 + x_2))))
}

r_hat <- function(x_0, x_1, x_2){
    n <- x_0 + x_1 + x_2
    l1 <- -log((x_0+x_1)/n)
    l2 <- -log(x_0/n)
    l1/l2
}

# plot (estimated) profile log-likelihoods for different values of counts
par(mfrow=c(3,2))
for(tt in list(list(x0=50, x1=10, x2=1),
            list(x0=50, x1=1, x2=10),
            list(x0=50, x1=25, x2=25),
            list(x0=500, x1=250, x2=250),
            list(x0=20000, x1=3, x2=3),
            list(x0=20000, x1=10, x2=2))){
    x0 <- tt[["x0"]]
    x1 <- tt[["x1"]]
    x2 <- tt[["x2"]]
    
    lam_ml <- -log(x0/(x0+x1+x2))

    rr <- seq(0.01,0.99, length.out=1000)
    ll0 <- lp(x0, x1, x2, rr)
    ll1 <- le(x0, x1, x2, rr)
    ll2 <- loglik_ind(x0, x1, x2, rr, lam_ml)
    mle <- r_hat(x0, x1, x2)
    mle_l1 <- loglik(x0, x1, x2, mle, lam_ml)
    mle_l2 <- loglik_ind(x0, x1, x2, mle, lam_ml)

    lower_e <- uniroot(function(x){2*(le(x0,x1,x2,x) - le(x0,x1,x2,r_hat(x0, x1, x2))) + qchisq(0.95, 1)},
                           c(0,r_hat(x0, x1, x2)))$root
    upper_e <- uniroot(function(x){2*(le(x0,x1,x2,x) - le(x0,x1,x2,r_hat(x0, x1, x2))) + qchisq(0.95, 1)},
                           c(r_hat(x0, x1, x2), 1))$root
    
    lower_bin <- uniroot(function(x){2*(loglik_ind(x0,x1,x2,x,lam_ml) - loglik_ind(x0,x1,x2,r_hat(x0, x1, x2),lam_ml)) + qchisq(0.95, 1)},
                           c(0,r_hat(x0, x1, x2)))$root
    upper_bin <- uniroot(function(x){2*(loglik_ind(x0,x1,x2,x,lam_ml) - loglik_ind(x0,x1,x2,r_hat(x0, x1, x2),lam_ml)) + qchisq(0.95, 1)},
                           c(r_hat(x0, x1, x2), 1.4))$root

    se2 <- se2_r_1(mle*lam_ml, lam_ml, x0+x1+x2, x0+x1+x2)


    plot(rr, ll1-max(ll1), type="l", col="red", 
        xlab=expression(r), ylab="relative profile log-likelihood", 
        main=paste("x0=",x0,", x1=",x1, " x2=",x2, sep=""))
    lines(rr, ll0-max(ll0), col="black")
    lines(rr, ll2-max(ll2), col="blue")
    abline(v=mle)
    
    rect(lower_e, min(ll1), upper_e, 10, col=rgb(red=1, green=0, blue=0, alpha=0.1), border=NA)
    rect(mle-1.96*sqrt(se2), min(ll1), mle+1.96*sqrt(se2), 10, col=rgb(red=0, green=1, blue=0, alpha=0.1), border=NA)
    rect(lower_bin, min(ll1), upper_bin, 10, col=rgb(red=0, green=0, blue=1, alpha=0.1), border=NA)
    
    yrange = range(ll1-max(ll1), finite=T)
    segments(lower_e, yrange[1]+0.5*diff(yrange), upper_e, yrange[1]+0.5*diff(yrange), col="red")
    segments(lower_bin, yrange[1]+0.4*diff(yrange), upper_bin, yrange[1]+0.4*diff(yrange), col="blue")
    segments(mle-1.96*sqrt(se2), yrange[1]+0.3*diff(yrange), mle+1.96*sqrt(se2), yrange[1]+0.3*diff(yrange), col="green")
    
    legend("bottomright",
           legend=c(paste("multinomial model: [", round(lower_e,2), ", ", round(upper_e,2), "]", sep=""),
                    paste("product of binomials: [", round(lower_bin,2), ", ", round(upper_bin,2), "]", sep=""),
                    paste("delta method: [", round(mle-1.96*sqrt(se2),2), ", ", round(mle+1.96*sqrt(se2),2), "]", sep="")),
       col=c("red", "blue", "green"), lty=c(1,1,1), cex=1,
          bg="white")
    



}
```

