---
title: "fitting_binomial_data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fitting_binomial_data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(WWdPCR)
```

# Example 1: simulated data

## Generate simulated data

Simulate data from a 3PL logistic model, with lower asymptote at 0.2, and binomial distribution of the data:
```{r example1}
set.seed(43)

N <- 30
xx1 <- 0:N
yy1 <- pl3model(xx1, 0.5, -10, 0.2)
yy2 <- rbinom(N+1, 20, yy1) / 20
```

## Fit model and inspect results

Fit a 3PL model with binomial noise:
```{r}
fit_3PL <- fit_pl(yy2, xx1, 20)
coef(fit_3PL)
confint(fit_3PL)

```

## Plot result:

```{r, fig.dim=c(6, 5)}
plot(xx1, yy2)
lines(xx1, yy1)
lines(xx1, fitted(fit_3PL), col="red")
legend("bottomright", legend=c("true", "3PL fit"),
       col=c("black", "red"),
       lty=c(1,1), cex=1)
```

## Plot confidence bands for fit 

We can predict, and compute standard errors either in the original linear scale, or in the logit scale. Computing standard errors in the logit scale allows for computing confidence intervals that, when back transformed to the linear scale, are guaranteed to be restricted to the [0,1] range.

```{r, fig.dim=c(7, 5)}
predicted <- predict(fit_3PL, scale="linear", se=TRUE) #returns a list with "pred" and "se" entries
plot(xx1, yy2)
lines(xx1, predicted$pred, col="black")

#plot confints calculated from linear space se
lines(xx1, predicted$pred - 1.96*predicted$se, lty=2, col="red")
lines(xx1, predicted$pred + 1.96*predicted$se, lty=2, col="red")

# #plot confints calculated from logit space se
predicted_logit <-  predict(fit_3PL, scale="logit", se=T)
lines(xx1, logit_inv(predicted_logit$pred - 1.96*predicted_logit$se), lty=2, col="blue")
lines(xx1, logit_inv(predicted_logit$pred + 1.96*predicted_logit$se), lty=2, col="blue")
legend("bottomright", legend=c("3PL fit", "linear confint", "logit confint"),
       col=c("black", "red", "blue"),
       lty=c(1,2,2), cex=1)

```


# Example 2: fitting on Swiss HV69-70del clinical data

## Load and prepare data

We have daily prevalence of the mutation in the sequencing results:

```{r}
str(clinical_HV6970del)
```

Let's prepare the data:

```{r}
# Make numerical values out of the dates, starting at october first 2020
clinical_del_6970_xx <- as.numeric(as.Date(clinical_HV6970del$date) - as.Date("2020-10-01 00:00:00"))
# Normalize counts and filter out values before October first
clinical_del_6970_yy <- (clinical_HV6970del$HV69.70del / clinical_HV6970del$sequenced)[clinical_del_6970_xx >= 0]
# Make a vector of sample sizes for weights and filter out values before October first
clinical_del_6970_ww <- (clinical_HV6970del$sequenced)[clinical_del_6970_xx >= 0]
# Filter out values before October first from the numerical dates
clinical_del_6970_xx <- clinical_del_6970_xx[clinical_del_6970_xx >= 0]
```

## Fit and inspect

Fit a 2PL model accounting for overdispersion. We need to play around with starting values a bit. 

```{r}
clinical_del6970_2PL <- fit_pl(clinical_del_6970_yy, clinical_del_6970_xx, clinical_del_6970_ww,
                               fitfun = pl2model,
                               quasi=T, 
                               starting_values = c("a"=0.1, "b"=-10), 
                               lower = c(0,-50, 0), upper = c(1, 50, 1))
confint(clinical_del6970_2PL)
```

Fit a 3PL model, we can take estimates of the 2PL as initial guesses for the concerned params, and some other "clever" estimates:

```{r}
clinical_del6970_3PL <- fit_pl(clinical_del_6970_yy, clinical_del_6970_xx, clinical_del_6970_ww,
                               quasi=T, 
                               starting_values = c("a"=0.05, "b"=-6.46, 
                                                   "c"=mean(clinical_del_6970_yy[1:10])), 
                               lower = c(0,-50, 0), upper = c(1, 50, 1))
confint(clinical_del6970_3PL)
```

## Test 3PL vs 2PL

Make a quasi log-likelihood ratio test:

```{r}
# compute quasi log-likelihood ratio, with inflation factor from 2PL (more conservative)
llr <- -2*((-1*clinical_del6970_3PL$likelihood/clinical_del6970_2PL$infl_factor) - (-1*clinical_del6970_2PL$likelihood/clinical_del6970_2PL$infl_factor))
# chisquare test of the log-likelihood ratio
print(c("2logLr:"=llr, "p-val:"=formatC(1-pchisq(llr, 1), format = "e", digits = 2)) )

```


## Plot

```{r, fig.dim=c(7, 5)}
# plot raw data and fits
plot(clinical_del_6970_xx, clinical_del_6970_yy,
     xlab="days since 2020-10-01",
     ylab="frac. HV69-70del in clinical samples")
lines(sort(clinical_del_6970_xx),
      clinical_del6970_3PL$fitted,
      col="red")
lines(sort(clinical_del_6970_xx),
      clinical_del6970_2PL$fitted,
      col="black")

# make confints through logit reparametrization and plot them
predicted_3PL_logit <-  predict(clinical_del6970_3PL, scale="logit", se=T)
polygon(c(clinical_del_6970_xx, rev(clinical_del_6970_xx)),
        c(logit_inv(predicted_3PL_logit$pred - 1.96*predicted_3PL_logit$se),
          rev(logit_inv(predicted_3PL_logit$pred + 1.96*predicted_3PL_logit$se))),
       col=rgb(red=1, green=0, blue=0, alpha=0.1), border=NA)
predicted_2PL_logit <-  predict(clinical_del6970_2PL, scale="logit", se=T)
polygon(c(clinical_del_6970_xx, rev(clinical_del_6970_xx)),
        c(logit_inv(predicted_2PL_logit$pred - 1.96*predicted_2PL_logit$se),
          rev(logit_inv(predicted_2PL_logit$pred + 1.96*predicted_2PL_logit$se))),
       col=rgb(red=0, green=0, blue=1, alpha=0.1), border=NA)

# make legend
legend("bottomright", legend=c("3PL fit", "2PL fit"),
       col=c("red", "black"), lty=c(1,1), cex=1)

```

## Inspect overdispersion:

It looks like a constant overdispersion factor assumption is ok, and that we don't want a beta binomial distribution (that would imply a linear dependency between overdispersion and sample size). 

```{r, fig.dim=c(7, 5)}
plot(clinical_del_6970_ww, log10(clinical_del6970_3PL$overdispersion),
    xlab="sample size", 
    ylab="log10 variance ratio", 
    main="overdispersion vs sample size")
for (i in 1:100){
    indxs <- sample(1:length(clinical_del_6970_ww), replace=T, prob=clinical_del_6970_ww)
    lines(loess.smooth(clinical_del_6970_ww[indxs], log10(clinical_del6970_3PL$overdispersion)[indxs]),
          col=rgb(red=1, green=0, blue=0, alpha=0.1))
}
lines(loess.smooth(clinical_del_6970_ww, log10(clinical_del6970_3PL$overdispersion)), col="red")
points(clinical_del_6970_ww, log10(clinical_del6970_3PL$overdispersion))
```



